---
title: "Mini R Project 2"
author: "Natalia Iannucci and Sunni Raleigh"
output: html_document
---

##loading the dataset
```{r}
data <- read.table("Shortleaf.txt", header = T, sep = '\t')

attach(data)

```

##Initial study to observe the nature of the relationship between the volume of shortleaf pines and their diameters 

__Graphical Method:__ Scatter Plot
```{r}
plot(Diameter, Volume, pch = 19, cex = 1.5, col = "blue", main = "Scatter plot between volume and diamater of shortleaf pines")
```

The scatterplot shows a positive trend, however; it appears to be slightly non-linear.

__Numerical Method:__ Pearson Linear Correlation Coefficient 
```{r}
r <- cor(Diameter, Volume)
r
```

The r value of 0.9447509 suggests that there is a strong positive linear correlation between tree diameter and volume.

__Based on the r value, fitting a linear relationship might be reasonable. However, the non-linear trend in the scatter plot suggests that a linear relationship may not be best, so we will fit a linear relationship and then assess whether or not it is a good fit. __ 

##Fit a least square regression line to predict the volume based on the diameter.
```{r}
regModel <- lm(Diameter ~ Volume)
regModel
```

$$y_i = 6.6285 +0.1306x_i + e_i$$ for $i=1,2,\cdots,70$ where y is the tree volume, x is the tree diameter, and e is the random error with the assumptions $e_i \stackrel{iid}\sim N(0, \sigma^2)$.

```{r}
plot(Diameter, Volume, pch = 16, main = "Fitted Regression Line")
abline(regModel, col= "blue", lwd = 2)

```

##Check Conditions:

###__Find Residuals and Predicted Values:__
```{r}
# find residuals
residuals <- resid(regModel)
residuals

# find predicted values
predValues <- predict(regModel)
predValues
```


###__Zero Mean:__
```{r}
m <- mean(residuals)
m
# the zero mean condition holds.
```
The mean of the residuals is very close to 0 and and the zero mean condition holds


###__Linearity and Constant Variance:__
```{r}
# plot the predicted values against the residuals
plot(predValues, residuals)
abline(0,0, col="red")
```
The condition of constant variance and linearity are not met, as the variance fans out and the scatterplot of residuals shows a clear nonlinear trend. This suggests that we need to perform a transformation.


###__Normality:__
```{r}
# Histogram
hist(residuals)

# Q-Q Plot
qqnorm(residuals, pch = 19, col= "blue")
qqline(residuals, col="red")

## Numerical Method
shapiro.test(residuals)
```
At a 5% level, the Shapiro-Wilk test suggests the residuals follow a normal distribution. However, the histogram of residuals appears to be left-skewed, and the points on the normal Q-Q plot are not linear. Numerically, normality holds but graphically it does not. Therefore, we have chosen to do a transformation. 

---

## Remedial Measures
For our transformation we have decided to transform Y because constant variance is one of the  one of the conditions violated in our first model. $$ transformation: y = \sqrt{y}$$
```{r}
sqrt_Y = sqrt(Volume)
r = cor(sqrt_Y, Diameter)
r

plot(Diameter, sqrt_Y, pch = 16, ylab = "Volume", xlab = "Diamater", main = "scatter plot between volume and diamater")
```
The scatterplot drawn between the transformed Y and X reflects a linear relationship, with a r value of 0.987 suggesting a strong linear relationship between sqrt(Y) and X modeled by the following:

```{r}
newRegModel <- lm(sqrt_Y~Diameter)
newRegModel
```
$$\sqrt{y_i} = -1.1366 + 0.5832x_i + e_i$$ for $i=1,2,\cdots,70$ where y is the tree volume, x is the tree diameter, and e is the random error with the assumptions $e_i \stackrel{iid}\sim N(0, \sigma^2)$.

## Plot new model
```{r}
plot(Diameter, sqrt_Y, pch = 16, main = "Fitted Regression Line")
abline(newRegModel, col= "blue", lwd = 2)
```

##Check conditions for new model:

###__Find new Residuals and new Predicted Values:__
```{r}
# find residuals
newResiduals <- resid(newRegModel)
newResiduals

# find predicted values
newPredValues <- predict(newRegModel)
newPredValues
```


###__Zero Mean:__
```{r}
m <- mean(newResiduals)
m
```

###__Linearity and Constant Variance:__
```{r}
# plot the predicted values against the residuals
plot(newPredValues, newResiduals)
abline(0,0, col="red")
```
The variance of the residuals is fairly constant and there is a slight linear trend. This shows considerable improvement from our original model. Constant variance and linearity hold.

## Check normality of new residual values:
```{r}
hist(newResiduals)

qqnorm(newResiduals, pch = 19, col= "blue")
qqline(newResiduals, col="red")

shapiro.test(newResiduals)
```
While the distribution of the new residuals is slightly skewed left, the points on the Q-Q line are linear and the results from the Shapiro-Wilk test at a 5% level suggest that the points follow a normal distribution therefore the condition for normality holds.
